|

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Make notes and style changes
(Several channges altogether, too many!)

---
title:     |
  A generalization of multidimensional item response theory parameters
author:    |
  <table>
    <tr>
      <td>Daniel Morillo, Ph.D.</td>
      <td>Mario Luzardo, Ph.D.</td>
    </tr>
    <tr>
      <td>
        <center>
          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/DaniMori/)&emsp;
          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0003-3021-3878)
        </center>
      </td>
      <td>
        <center>
          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/mluzardov/)&emsp;
          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0002-9360-2806)
        </center>
      </td>
    </tr>
  </table>
institute: |
  <table>
    <!-- # TODO: Fix vertical alignment in logos-->
    <tr>
      <td style="width: 400px">
        [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="60"}](https://www.uned.es/universidad/facultades/psicologia.html)
      </td>
      <td style="width: 100px">
        [![](../www/slideshow-assets/logo_udelar.svg){height="120"}](https://udelar.edu.uy)
      </td>
    </tr>
  </table>
bibliography:  ../www/Multidimensional-parameters-MCLM.bib
csl:           ../www/apa-old-doi-prefix.csl
date:          "2024-07-17"
date-meta:     "2024-11-27"
date-format:   long
editor:
  mode:        source
  markdown: 
    wrap:      sentence
    canonical: true # TODO: Restore
execute: 
  cache:       true
lang:          en
knitr:
  # opts_knit: # Does not work, see line 102
  #   root_dir:  here::here()
  opts_chunk: 
    results:   asis
    autodep:   true
format:
  revealjs:
    auto-stretch:            true
    code-annotations:        hover
    df-print:                paged
    fig-cap-location:        bottom
    fig-format:              svg
    fig-asp:                 1
    fig-width:               6
    incremental:             false
    keep-md:                 false
    link-external-newwindow: true
    margin:                  0
    self-contained:          true
    slide-number:            false
    theme:                   ../www/slideshow-assets/extra-styles.scss
    transition:              none
    view-distance:           3
    template-partials:
      - ../www/slideshow-assets/title-slide.html
editor_options:
  chunk_output_type: console
---

@@

@@

::: notes
The Multidimensional 2PL model is one of many extensions of the 2PL model.

It assumes that the probability of a correct response to an item is a function of the linear combination of the latent traits weighted by item discrimination parameters.
(This here is a vector inner product; the "T" means that the discrimination vector is transposed).

And it has this additional "intercept parameter"; so it is basically a logistic multivariate regression with latent predictors.

@@

::: notes
The problem with the M2PL is that its parameters do not have a clear interpretation in IRT terms.

@@

$$
  MID_i = \begin{Bmatrix}
    D_i &= \frac{-d_i}{MDISC_i} \\
    \cos \mathbf{\alpha}_i &= \frac{\mathbf{a}_i}{MDISC_i}
  \end{Bmatrix}
$$
:::

::: {.column .fragment}
-   @ackerman_graphical_1996:

@@

::: notes
...Reckase & McKinley define the item discrimination, which is an extension of the Pythagorean theorem.

This discrimination can be used to define the item difficulty, finding a parameterization that parallels the unidimensional 2PL model.

Note though that, in this case, the difficulty has two components: a "signed distance" from the origin, and a direction.

It is easier to appreciate this with the graphical representation, that was proposed by Ackerman a few years later, where we can see that the item vector is applied at a point given by the difficulty (with its distance and direction), and has a length equal to the discrimination.
The vector direction always goes through the origin, and the sign of the distance (positive in this case), which is the opposite of the intercept sign, will determine whether the vector is displaced forward or backward along that direction.
:::

@@

::: notes
However, there are two assumptions for defining the multidimensional parameters.

First, that the probability of a correct answer increases monotonically with each dimension.

And second, this constrain you see there, which implies that the item is represented in an orthogonal basis.
This is necessary for applying the "general Pythagorean theorem" as we saw before.

So, our purpose is **to relax these two assumptions**.
:::

@@

::: {.column .fragment fragment-index="1"}
```{r inverse-item-curve}
#| fig-align:  center
#| fig-asp:    0.5344
#| fig-width:  9

@@

::::: {.columns .fragment fragment-index="2"}
::: column
$$
  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
$$
:::

@@

::: {.fragment .r-center}
### Generalization: "Constant monotony"
:::

::: notes
Because they derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to non-cognitive items as well, e.g. to personality measures.

@@

Here, we can have items with "decreasing" probability.

Formally, what this would imply is a negative discrimination parameter.

In the multidimensional case, this may imply a mix of positive and negative discrimination parameters, even in the same item.
:::

@@

::::::: {.fragment fragment-index="1"}
::: {style="color:#5c6eb1; padding-left:300px;"}
$d_1 = 0$
:::

::: {style="color:#90214a; padding-left:300px;"}
$d_2 = -0.5$
:::

::: {style="color:#749f4c; padding-left:300px;"}
$d_3 = -0.25$
:::

::: {style="color:#d76f47; padding-left:300px;"}
$d_4 = 0.25$
:::
:::::::
::::::::

@@

::: notes
The generalization is straightforward: The MDISC implies squared values of the discrimination components (this is the inner product expanded), so it is always positive regardless of the sign of the discrimination components.

The signed distance is also the same, and the direction only needs to take into account the sign of each component when computing that corresponding cosine.

@@

Graphically, we can see that each item points in a direction given by its discrimination components, including their signs (so, 1 1, 1 -1, etc.).

@@

This seems obvious enough, but we must explicitly state that this definitions are valid in these cases, and that they have the same interpretation.

@@

::::: columns
::: column
```{r item-oblique-axes-4}
#| fig-align:  center
#| fig-height: 8
#| fig-width:  8
mirt_item_transf <- tribble(
  ~item, ~a_1, ~a_2, ~d,
  1,     1,    1,    0
)

@@

:::::::::: columns
:::::: column
$$
  cor (\mathbf{\theta}) = \mathbf{\Sigma}
$$

@@

::: {.fragment fragment-index="2"}
$$
  cor (\mathbf{\theta}^*) =
    \mathbf{\Sigma}^* =
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
$$
:::
::::::

@@

::: notes
So, let's say $\theta$ is multivariate normal with covariane matrix $\mathbf{\Sigma}$.

@@

And if we want the geometrical space to be a representation of the latent space, the meaningful way is to represent these coordinates in an orthonormal basis (the covariane matrix is an identity matrix, the basis also forms an identity matrix).
:::

@@

thetas_transf |>
  ggplot(mapping = aes(trait_1, trait_2)) +
  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
  scale_x_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL,
    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
  ) +
  scale_y_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL
  ) +
  coord_fixed() +
  theme(
    axis.ticks   = element_blank(),
    axis.line    = element_blank(),
    axis.title   = element_markdown(),
    axis.title.y = element_markdown(),
    panel.grid   = element_blank()
  )
```
:::

::: {.column .fragment width="33%" fragment-index="1"}
```{r question-mark}
fa(
  "question",
  fill           = "#929acb",
  height         = "200px",
  vertical_align = "-10em",
  margin_left    = "4em"
)
```
:::

::: {.column .fragment width="33%" fragment-index="2"}
```{r corr-traits-transf-example}
#| fig-width:  5.5
#| fig-height: 5.5
```
:::
::::::

@@

::: {.fragment fragment-index="2"}
Which coordinate system *θ* makes *θ\** "spherical"?
:::

@@

In other words,...

...what is the coordinate system of $\mathbf{\theta}$...

@@

@@

Take a few seconds to answer and we will see next which is the correct one.
:::

@@

::: notes
As it turns out, this gives a space where the axes actually do represent faithfully the latent covariances.
:::

@@

@@

::: fragment
> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by $`r INNER_PROD_EQ`$.

@@

::: notes
This is actually not new, it was already noted by Zhang & Stout, more than 25 years ago, to solve a different problem, but unfortunately they didn't provide a deduction as we do here.

But the important thing here, is that we have two different spaces: One for the latent traits and one for the items.

@@

## Generalized MIRT parameters {.smaller}

@@

$$
`r latex_eq(
     MIL_COV_PARAM,
     latex_curlybraces("$DISTANCE_COV_DEF$, $DIR_COS_ITEM_VEC_COV_DEF$")
   )`
$$
:::

::: column
```{r item-plot-out}
#| fig-cap:   Item representation in non-orthonormal test space
#| fig-align: center
plot_oblique +
  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
  theme(axis.ticks = element_blank()) ## TODO: Tema ggplot
```
:::
:::::

@@

\[IF THERE IS TIME\]

@@

# The end

<!-- TODO: Add contact info -->

@@

>>>>

---
title:     |
  A generalization of multidimensional item response theory parameters
author:    |
  <table>
    <tr>
      <td>Daniel Morillo, Ph.D.</td>
      <td>Mario Luzardo, Ph.D.</td>
    </tr>
    <tr>
      <td>
        <center>
          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/DaniMori/)&emsp;
          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0003-3021-3878)
        </center>
      </td>
      <td>
        <center>
          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/mluzardov/)&emsp;
          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0002-9360-2806)
        </center>
      </td>
    </tr>
  </table>
institute: |
  <table>
    <!-- # TODO: Fix vertical alignment in logos-->
    <tr>
      <td style="width: 400px">
        [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="60"}](https://www.uned.es/universidad/facultades/psicologia.html)
      </td>
      <td style="width: 100px">
        [![](../www/slideshow-assets/logo_udelar.svg){height="120"}](https://udelar.edu.uy)
      </td>
    </tr>
  </table>
bibliography:  ../www/Multidimensional-parameters-MCLM.bib
csl:           ../www/apa-old-doi-prefix.csl
date:          "2024-07-17"
date-meta:     "2024-11-27"
date-format:   long
editor:
  mode:        source
  markdown: 
    wrap:      sentence
    canonical: true # TODO: Restore
execute: 
  cache:       true
lang:          en
knitr:
  # opts_knit: # Does not work, see line 102
  #   root_dir:  here::here()
  opts_chunk: 
    results:   asis
    autodep:   true
    cache:     true
format:
  revealjs:
    auto-stretch:            true
    code-annotations:        hover
    df-print:                paged
    fig-cap-location:        bottom
    fig-format:              svg
    fig-asp:                 1
    fig-width:               6
    footer:                  https://bit.ly/vote-coords
    incremental:             false
    keep-md:                 false
    link-external-newwindow: true
    margin:                  0
    self-contained:          true
    slide-number:            false
    smaller:                 true
    theme:                   ../www/slideshow-assets/extra-styles.scss
    transition:              none
    view-distance:           3
    template-partials:
      - ../www/slideshow-assets/title-slide.html
editor_options:
  chunk_output_type: console
---

@@

## Participate!

![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="400px"}

<https://bit.ly/vote-coords>

@@

::: notes
The Multidimensional 2-Parameter Logistic model (M2PL) is one of many extensions of the 2PL model.

It assumes that the probability of a correct response to an item is a function of the linear combination of the latent traits weighted by item discrimination parameters.

@@

::: notes
The M2PL has a problem though, in that its parameters do not have a clear interpretation in IRT terms.

@@

$$
  MIL_i = \begin{Bmatrix}
    D_i &= \frac{-d_i}{MDISC_i} \\
    \cos \mathbf{\alpha}_i &= \frac{\mathbf{a}_i}{MDISC_i}
  \end{Bmatrix}
$$
:::

::: column
-   @ackerman_graphical_1996:

@@

::: notes
...Reckase & McKinley define the item discrimination, which you can see is like a multidimensional version Pythagoras' theorem.

This MDISC can be used to define the item difficulty (or location as we will call it here), finding a parameterization that parallels the unidimensional case.

Only that, in this case, the location has two components: a "signed distance" from the origin, which is the opposite of the intercept sign, and a direction.

The item vector is applied at a point given by the location (with its distance and direction), and has a length equal to the discrimination.
The direction always goes through the origin, and the sign of the distance (positive in this case), will determine whether the vector is displaced forward or backward along that direction.
:::

@@

::: notes
However, there are two assumptions for defining these parameters.

First, that the probability of a positive response increases monotonically with each dimension.

And second, this constrain you see there, which implies that the item is represented in orthogonal coodrinates.
This is necessary for applying the "multidimensional Pythagoras' theorem" as we saw before.

So, our purpose is **to relax these two assumptions**, which derive heavily from the traditional application of this model to the cognitive domain.

But we want to apply multidimensional IRT to non-cognitive items as well, like personality measures, for example.
:::

@@

::: column
```{r inverse-item-curve}
#| fig-align:  center
#| fig-asp:    0.5344
#| fig-width:  9

@@

::::: {.columns .fragment}
::: column
$$
  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
$$
:::

@@

::: fragment
### Generalization: "Constant monotony"
:::

::: notes
In these cases, we can have items with "decreasing" probability.

@@

Which formally implies a negative discrimination parameter.

In the multidimensional case, we may have a mix of positive and negative discrimination parameters.
:::

@@

::::::: {.fragment fragment-index="1"}
::: {style="color:#5c6eb1; padding-left:300px;"}
$D_1 = 0$
:::

::: {style="color:#90214a; padding-left:300px;"}
$D_2 = 0.35$
:::

::: {style="color:#749f4c; padding-left:300px;"}
$D_3 = 0.18$
:::

::: {style="color:#d76f47; padding-left:300px;"}
$D_4 = -0.18$
:::
:::::::
::::::::

@@

::: notes
The generalization is straightforward: The MDISC implies squared values of the discrimination components (this is the previous inner product expanded), so it is always positive.

The signed distance is the same, and the direction only needs to take into account one sign, that of the corresponding component for computing each cosine.

@@

We can see that each item points in a direction given by its discrimination components with their signs (so, 1 1, 1 -1, etc.).

@@

This seems obvious enough, but we must explicitly state that this definitions are valid in these cases, and that their interpretation do not change.

@@

::::: columns
::: column
```{r item-oblique-axes-3}
#| fig-align:  center
#| fig-height: 8
#| fig-width:  8
mirt_item_transf <- tribble(
  ~item, ~a_1, ~a_2, ~d,
  1,     1,    1,    0
)

@@

:::::::::: columns
:::::: column
$$
  cov (\mathbf{\theta}) = \mathbf{\Sigma}
$$

@@

::: {.fragment fragment-index="2"}
$$
  cov (\mathbf{\theta}^*) =
    \mathbf{\Sigma}^* =
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
$$
:::
::::::

@@

::: notes
So, let's say our latent trait vector is multivariate normal with covariane matrix $\mathbf{\Sigma}$.

@@

And if we want the geometrical space to be a representation of the latent space, the meaningful way is to represent these coordinates in an orthonormal basis (the covariance matrix is an identity matrix, the basis of this coordinate system also forms an identity matrix).

But, if this is the coordinate system of $\mathbf{\theta^*}$, if we want to draw here the axes of $\mathbf{\theta}$, they cannot be these same ones, right?
If they were, then the point cloud would be the ellipse we had before.
:::

@@

thetas_transf |>
  ggplot(mapping = aes(trait_1, trait_2)) +
  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
  scale_x_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL,
    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~} ?", labels = NULL)
  ) +
  scale_y_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL
  ) +
  coord_fixed() +
  theme(
    axis.ticks   = element_blank(),
    axis.line    = element_blank(),
    axis.title   = element_markdown(),
    axis.title.y = element_markdown(),
    panel.grid   = element_blank()
  )
```
:::

::: {.column .fragment width="33%" fragment-index="1"}
```{r question-mark}
fa(
  "arrow-right",
  fill           = "#929acb",
  height         = "200px",
  vertical_align = "-9.5em",
  margin_left    = "2em"
)
```
:::

::: {.column .fragment width="33%" fragment-index="1"}
```{r corr-traits-transf-example}
#| fig-width:  5.5
#| fig-height: 5.5
```
:::
::::::

@@

@@

In other words: What is the coordinate system of $\mathbf{\theta}$...

@@

<center>$$
  \mathbf{\Sigma} = \begin{bmatrix}
      1 & 0.5 \\
      0.5 & 1
    \end{bmatrix}
$$</center>

Remember that in this example, the covariance matrix has unitary variance and .5 correlation.
:::

![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="400px"}

<https://bit.ly/vote-coords>

:::::::::::::::::::::::::::::::::::::::: columns
::::::::::::::::::::::::::::::::::::::: {.column width="33%"}
<center>

A?`<`{=html}
`/center>`{=html}

```{r grid-orthogonal-rotation-spherical-option}
#| fig-width: 5
rot_basis <- cbind(
  c(sqrt(.75), -sqrt(.25)),
  c(sqrt(.25), sqrt(.75))
)

rot_grid +
    geom_point(
    mapping = aes(trait_1, trait_2),
    data    = thetas_transf,
    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
  )
```

:::

:::::::::::::::::::::::::::::::::::::: {.column width="33%"}
<center>

B?`<`{=html}
`/center>`{=html}

```{r grid-ts-basis-spherical-option}
#| fig-width: 5
ts_grid <- transform_grid(ts_basis)

ts_grid +
    geom_point(
    mapping = aes(trait_1, trait_2),
    data    = thetas_transf,
    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
  )
```

:::

::::::::::::::::::::::::::::::::::::: {.column width="33%"}
<center>

C?`<`{=html}
`/center>`{=html}

```{r grid-ls-basis-spherical-option}
#| fig-width: 5
ls_grid <- transform_grid(ls_basis)

ls_grid +
    geom_point(
    mapping = aes(trait_1, trait_2),
    data    = thetas_transf,
    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
  )
```

:::

::::::

@@

::: notes
\[CLICK LINK AFTER A WHILE\]
:::

@@

::: notes
As it turns out, this gives a space where the axes actually do represent faithfully the latent covariances.

But it is not the latent space, where the latent traits are represented.
It is a specific space for the items, and we call it "test space".
And we know it, because the inner products in both spaces have to be different from each other, for the model to be invariant to changes of bases, like the one you have just seen.
:::

@@

> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by $`r INNER_PROD_EQ`$.

--- @zhang_theoretical_1999 [p. 221]

@@

@@

::: notes
This is actually not new, it was already noted by Zhang & Stout, more than 25 years ago, to solve a different problem, but unfortunately they didn't provide a mathematical proof as we do here.

As you see here, the latent space Gram matrix is the inverse covariance matrix (which results from the transformation matrix P, pre-multiplied by itself, P\^T P).
The invariant distance then, in the latent space, is the Mahalanobis distance, which is the square root of the inner product of the latent trait vector with itself.
And in the test space, the Gram matrix is the covariance distance.

But the important thing here, is that we have two different spaces, and that the item representation must be in their own space, the test space.

@@

## Generalized M2PL parameters {.smaller}

@@

$$
`r latex_eq(
     MIL_COV_PARAM,
     latex_curlybraces("$DISTANCE_COV_DEF$, $DIR_COS_ITEM_VEC_COV_DEF$")
   )`
$$

$$
  \mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}
$$
:::

::: {.column .fragment}
```{r item-plot-out}
#| fig-cap:   Item representation in non-orthonormal test space
#| fig-align: center
plot_oblique +
  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
  theme(axis.ticks = element_blank()) ## TODO: Tema ggplot
```
:::
:::::

@@

The MDISC is the square root of the inner product of the discrimination vector; the signed distance does not change.

And the direction cosines incorporate this term: It pre-multiplies the discrimination vector by the covariance matrix, and also has this "normalizing term" that is a diagonal matrix of standard deviations..

\[IF THERE IS TIME; NEXT\]

@@

@@

::: {#refs}
:::

## The end

### Thanks for your attention!

::::: columns
::: column
![](../www/slideshow-assets/foto_daniel.jpg){width="200px" fig-align="center"}

<center>Daniel Morillo, Ph.D.</center>
:::

::: column
![](../www/slideshow-assets/foto_mario.jpg){width="200px" fig-align="center"}

<center>Mario Luzardo, Ph.D.</center>
:::
:::::

<center>Contact: [danielmorillo.ac\@gmail.com](mailto:danielmorillo.ac@gmail.com){.email}</center>
:::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::
:::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::
